%!TEX root=thesis.tex
\documentclass[thesis.tex]{subfiles}

\begin{document}

In the following Sections I give the context and main results for each of the four projects I am part of. 

The main project is the development of an effective equation of motion for the orientation of a neutrally buoyant spheroid suspended in a simple shear flow, valid when inertial effects are weak but not vanishing. In short, we compute what happens to the Jeffery orbits when the particle Reynolds number is non-zero. The results are contained in the appended Papers A-D.

In addition I have taken part in three studies related to the orientational motion of non-spherical particles. They have in common that they involve Jeffery's theory for ellipsoidal particles.
\begin{enumerate}
    \item a microfluidic experiment aiming to validate Jeffery's theory for the rotation of triaxial particles in shear flow (Paper E),
    \item a study of the effects of thermal noise on the orientational dynamics of triaxial particles in shear flow, with implications for the dilute suspension rheology (Paper F),
    \item a description of the rotational modes of disks and rods in isotropic turbulence, combining data from experiments, direct numerical simulations and random flow theory (Paper G).
\end{enumerate}


\chapter{Effective equation of motion for a weakly inertial spheroid in shear flow}

This project is a collaboration with collegues in Cherbourg and Marseille (France), and Stockholm (Sweden). J.R. Angilella (Cherbourg) and F. Candelier (Marseille) have many years of experience in dynamical systems, fluid mechanics and perturbation theory, without which this project would not have landed. T. Ros\'en and F. Lundell in Stockholm are experts in direct numerical simulation of particulate flows by the lattice Boltzmann method, by which we could validate our calculations. I also attribute the initial idea to perform stability analysis on the log-rolling motion under inertial perturbation, however vague at the time, to F. Lundell at a COST meeting in Udine, Italy. 

Paper A is a brief summary of the calculation and result in letter form, while the Papers B \& C contain all details. Paper D describes the direct numerical simulations which validates our theoretical calculation, and shows in detail when the effective equations break down due to finite domain size and increasing importance of inertial effects.

\section{History of problem}

The Jeffery equations for an axisymmetric particle in simple shear flow are interesting because their solutions are degenerate, as explained in \Secref{jefferyorbits} in Part I. Their solutions form a one-parameter family of periodic orbits. No orbit is preferred over another, it is the initial condition which determines the dynamics indefinitely. Jeffery was very aware of this fact, he writes \blockquote{It is obviously undesirable to leave a problem, which is physically quite determinate, in this indeterminate form.}
He further conjectures that \blockquote{this failure is due to the limitations of the theory of the slow motion of a viscous fluid.}
In other words, he believed that inertial corrections were necessary to break the degenaracy. He concludes, referring to inertial corrections, \blockquote[][...]{[...] a more complete investigation would reveal the fact that the particles do tend to adopt special orientations} 
In connection with this discussion Jeffery hypothesized that this preferred orientation would be such that the energy dissipated by viscosity is minimized: a prolate particle ends up log-rolling (long axis along the vorticity), and an oblate particle tumbles (with a diameter along vorticity.)

\citet{saffman1956} made the first attempt to include the non-linear inertial terms. It seems, although details are sparse, that he used an early form of asymptotic matching. For this he acknowledges I. Proudman, who a year later co-authored a paper \cite{proudman1957} on the inertial correction to the drag on a translating sphere, pioneering the use of asymptotic matching in viscous fluid mechanics. But Saffman did not have the proper solution to the outer ``Oseen problem'' for matching, instead he invented a plausible but ad-hoc boundary condition to match the inner expansion. With this method, applied for nearly spherical particles, he found agreement with Jeffery's minimum dissipation hypothesis.

\citet{harper1968} analyzed the rotation of two spheres rigidly constrained by an invisible rod, a so-called dumbbell. In the purely viscous regime a dumbell is equivalent to a prolate spheroid when its aspect ratio approaches infinity. They assumed that both spheres experienced lift forces (calculated by \citet{saffman1965}) independently of each other. By this method they found the opposite of Jeffery's minumum dissipation hypothesis: a slender rod ends up tumbling end-to-end in the flow-shear plane.

More recently \citet{subramanian2005,subramanian2006} re-examined the slender fibre and nearly-spherical limiting cases using a reciprocal theorem \cite{kim1991,lovalenti1993}. Their method is less controversial than those employed by the earlier attemps. However, they found, like \citet{harper1968}, that for small values of the Reynolds number a slender rod tumbles end-to-end in the flow-shear plane. For larger values of the Reynolds number the tumbling orbit is destroyed and replaced by fixed points, which means that the particle stops rotating and aligns, a phenomenen observed in numerical studies \cite{ding2000} (see below). \citet{subramanian2006} found that a nearly spherical prolate particle aligns its long axis along the vorticity, and a nearly spherical oblate particle tumbles, in agreement with \citet{saffman1956}. They remark that the different types of motions for nearly spherical particles, and slender particles \blockquote{suggests a possible bifurcation [...] at an intermediate aspect ratio.}

Meanwhile, several groups began studies of this problem using direct numerical solution of the Navier-Stokes equations, via lattice Boltzmann simulations \cite{feng1995,ding2000,qi2003,yu2007,huang2012,rosen2014,mao2014,rosen2015a,rosen2015b}. These studies reveal a rich structure of dynamical modes for moderate to large values of the Reynolds number. Not only log-rolling or tumbling is possible, but also intermediate limit cycles and alignment with new fixed points. The studies are limited to a few particle shapes, typically $\lambda=1/4$, $\lambda=1/2$, $\lambda=2$ and $\lambda=4$. Instead they focus on the effects of increasing Reynolds numbers, confinement and particle buoyancy.

I enter this chronology sometime in 2013, after submitting a paper \cite{einarsson2014} in which we describe the effects of particle inertia alone, neglecting the fluid inertia. We were intrigued by the results of \citet{subramanian2006,subramanian2006} outlined above, and the fact that no numerical results had shown the predicted log-rolling mode for nearly spherical prolate particles. For example, \citet{qi2003} simulated both oblate ($\lambda=1/2$) and prolate ($\lambda=2$) spheroids, and found for that the oblate particle logs-rolls while the prolate particle tumbles, the opposite of the existing theoretical prediction. But, as \citet{subramanian2006} points out, there were several possible explanations for this discrepancy. First, the simulations were performed at moderately small Reynolds numbers, but not much smaller than unity, where a perturbation theory should be valid. Second, they were performed in a finite computational domain, whereas the theory is valid for an unbounded shear flow. Third, the particle aspect ratio $\lambda=2$ is not close to unity, whereas the theory assumed $\lambda\approx1$. The parameter ranges where the existing theory should be valid is also where the numerical simulations become computationally impractical. More precisely, small values of the Reynolds number, large distance to the boundaries, and extreme particle shape all add to the computational cost. Therefore any comparisons to existing theory were qualitative.

These facts convinced us to attempt relaxing the assumption of $\lambda\approx1$, and find the exact value of $\lambda$ for the cross-over from log-rolling to tumbling.

During my work several more numerical studies have appeared \cite{rosen2014,mao2014,rosen2015a,rosen2015b}. Despite improved methods and more raw computer power, there was still no evidence of log-rolling prolate spheroids at small values of the Reynolds number. The clearest example is in Fig.~12 of \citet{mao2014}, where they find that the dynamics of a nearly spherical particle ($\lambda=1.2$ and $\lambda=0.8$) also contradicts the theoretical prediction. Their belief is that this \blockquote{may be caused by the influence
of the higher-order effects...}, implying that the value of the Reynolds number in their simulation was out-of-range for the perturbation theory.

\section{Results}

We initially set out to calculate only the linear stability exponents of the log-rolling position. But it turned out that we could compute an explicit correction to Jeffery's equation of motion, which is more useful. Let $\ve n$ be the unit vector along the symmetry axis of the particle, and $\ma O$ and $\ma S$ the antisymmetric and symmetric parts of the flow gradient (see \Secref{forces} in Part I for details.) Then the result is
\begin{align}
\eqnlab{ndoteffective}
  \dot {\ve n} &=  
\ma P\left[\ma O \ve n + \Lambda\ma S\ve n\right]\\
&\,+\Reys\ma P\left[b_1 (\ve n \cdot \ma S \ve n)\ma S \ve n
+ b_2 (\ve n \cdot \ma S \ve n)\ma O \ve n 
+b_3\,  \ma O \ma S \ve n 
+ b_4\,  \ma S \ma S \ve n\right]\,. \nn
\end{align}
Here the first row is the result of \citet{jeffery1922}. The projection matrix $\ma P=\ma 1 - \ve n \ve n\transpose$ removes any component of the vector field which is not tangent to the sphere (see also \Secref{jefferyequation} in Part I.) The scalar parameters $\Lambda$ and $b_\alpha$ are determined by the particle aspect ratio $\lambda$. The shape factor $\Lambda=(\lambda^2-1)/(\lambda^2+1)$ was computed by Jeffery. Our main accomplishment is the calculation of $b_\alpha(\lambda)$.

The result \eqnref{ndoteffective} resolves the inconsistensies between earlier theories, and between theory and numerical simulations. It also conclusively refutes Jeffery's minimum dissipation hypothesis. I summarise these findings in the following Sections.

\subsection*{First effects of inertia}

The solution to Jeffery's equation, \Eqnref{ndoteffective} with $\Reys=0$, are the degenerate periodic Jeffery orbits. But just as Jeffery conjectured, an arbitrarily small amount of inertia breaks the degeneracy, and the particle adopts a preferred orbit. What the preferred orbit is depends on the particle aspect ratio, and initial condition as follows.
\begin{itemize}
    \item $0 < \lambda < \lambda_c \approx 1/7.3$: The particle log-rolls if released closed to log-rolling, and tumbles if released close to the tumbling orbit. The two regions are separated by an unstable intermediate orbit.
    \item $\lambda_c < \lambda < 0$: The particle log-rolls, whatever the initial condition.
    \item $0 < \lambda$: The particle tumbles, whatever the initial condition.
\end{itemize}

Our result \eqnref{ndoteffective} agrees with those of \citet{subramanian2005} in the limit $\lambda\to\infty$ (up to a factor of $8\pi$). However, we find that the earlier results valid for nearly spherical particles \cite{saffman1956,subramanian2006} are mistaken. This is a bold statement, but we took great care in deriving and checking our result. First, we asked F. Candelier to independently analyse the case of nearly spherical particles. We knew that my general solution must match his solution for nearly spherical particles exactly, as $\lambda\to1$. By comparing notes we eliminated mistakes until, very suddenly, our solutions matched. At that point we compared my solution to \citet{subramanian2005} as $\lambda\to\infty$, and found agreement up to a numerical factor of $8\pi$. We have not identified exactly where in their calculation this factor appears, but the likely culprit is in the definition of Green's functions for Stokes flow (see App.~A.2 in Paper~C.) With these comparisons we were confident enough to submit our calculations for review and publication. Second, we asked T. Ros\'en and F. Lundell to use their numerical codes to verify our result. Our effective equation \eqnref{ndoteffective} agrees very well with the full numerical solution as $\Reys\to0$ and the computational box size becomes large. This comparison is in Paper~D, see in particular \obs{Fig.~2}. We conclude that our effective equation is correct, which resolves the indeterminacy discovered by Jeffery in 1922. But we also see that the orientational dynamics of a non-spherical particle in shear flow is sensitive to both confinement (wall-effects), and to higher-order corrections in the Reynolds number.

\subsection*{Phase diagram for the effective equation}
In the previous Section I discussed the first effects of inertia, for which a perturbative effective equation like \Eqnref{ndoteffective} should be valid. For larger values of $\Reys$ it is not as clear-cut, we cannot be certain that the dynamics of the effective equation reflects the dynamics of the exact equations. Nevertheless it can be fruitful to examine the dynamics of the effective equations.

Numerical simulation of the exact equations reveals many different modes of rotation, depending on parameters such as particle aspect ratio, Reynolds number, confinement ratio and particle buoyancy. Log-rolling and tumbling are two such modes which exist already when $\Reys\approx0$. But for larger values of $\Reys$ the particles may cease to rotate and align, for example. Such fixed points of the dynamics also appear for the effective equation of motion \eqnref{ndoteffective} as $\Reys$ increases. The effective equation has two parameters, $\Reys$ and $\lambda$. In \Figref{phasediagram} I show a phase diagram of the different regions in this parameter space. Each region corresponds to a distinct mode of rotation, depending on which periodic orbits and fixed points exist in \Eqnref{ndoteffective} at those values of the parameters. The borders between regions are \emph{bifurcation lines}, where the qualitative dynamics change.

It is clear that the effective equation has no quantitative predictive power except for very small values of $\Reys$. It can, for example, in general not predict at which value of $\Reys$ a disk-shaped particle with $\lambda=1/12$ will cease rotating. But if a bifurcation line extends towards $\Reys=0$ somewhere, we know that it must \emph{connect} to a corresponding bifurcation line in the exact dynamics. This does not necessarily predict what will happen in the exact dynamics, but it does constrain the options, and may serve as a guide for further numerical analysis.

In the phase diagram three lines extend to $\Reys=0$. One of them separates the regions of stable and unstable tumbling for oblate particles, referred to in the previous Section. It connects to $\Reys=0$ approximately when $\lambda=1/7.3$. In Paper~D there is numerical evidence that this happens also in the exact equations. The two other bifurcation lines reach $\Reys=0$ only asymptotically as $\lambda\to0$ and $\lambda\to\infty$, corresponding to infinitely thin disks or rods.

One may hope that the effective equation is qualitatively correct in predicting what the first bifurcation is as $\Reys$ increases. The data of \citet{rosen2015b} and Paper~D indicate that the ``critical point'', where several bifurcation lines merge, does exist also in the exact dynamics. However, the bifurcation lines seem to be sensitive to the confinement ratio in the numerical simulations and as of now we do not have enough data to confirm nor refute any claims on equivalence.

\chapter{Measurements of asymmetric rods tumbling in microchannel flow}

This project is a collaboration with our experimentalist collegues in Gothenburg, Sweden. Roughly, the division of work is that they build and perform the experiment, and we design the specifications and perform the analysis.

The project was initiated with an intention of observing the scattering between Jeffery orbits due to thermal noise. But first we needed to observe plain Jeffery orbits, as a baseline. With hindsight that was naive, given the long list of skilled experimentalists before us, Sec.~II in Paper E is a more or less exhaustive list. Our focus has shifted away from thermal noise, to the sensitive dependence on particle shape and initial condition. In \Secref{triaxialintro} in Part I, I explained how Jeffery's result implies complicated dynamics for triaxial particles, even if the deviation from axisymmetry very small. We want to observe this effect in experiment.

We previously published two papers \cite{mishra2012,einarsson2013} describing our methods and some initial observations. In Paper~E we describe our most recent measurements. There are two main differences from before. First, we employ an optical trap to arrange particles for the experimental runs. This allows us to use the same particle several times, and to control its initial condition. Second, we have particles made from glass fibres which have a very symmetric cross-section.

In Paper~E we claim to observe both quasi-periodic and chaotic trajectories, for the same particle. 
\begin{enumerate}
    \item Pressure reversals revert the orientational dynamics, which excludes any non-reversible effects such as inertia or thermal noise,
    \item Several trajectories for the same particle, but with different initial conditions.
\end{enumerate}

\chapter{Effects of thermal noise on the tumbling of triaxial ellipsoids}
Paper F
\section{History of problem}
\section{Results}



\chapter{Rotation rates of particles in turbulence}
Paper~G started as the synthesis of discussions we had during a workshop at {\sc nordita} in Stockholm. For those I am grateful especially to E.~Variano and G.~Voth. The paper is a discussion of the rotations of axisymmetric particles in isotropic turbulence. I think the strength of this paper is it's breadth, as it contains pieces of experimental results, numerical results and analytical model calculations.

From what I remember the discussions started because of confusion between \emph{rotation rate} and \emph{tumbling rate}. In this context rotation rate means magnitude of the angular velocity: $|\ve \omega|$. The tumbling rate is the rate at which the symmetry axis of the particle turns: $|\dot{\ve n}|=|\ve \omega \cross \ve n|$. The two are of course kinematically related, because
\begin{align}
    |\ve \omega|^2 &= |\dot{\ve n}|^2 + |\ve \omega \cdot \ve n|^2\,.
\end{align}
The difference $|\ve \omega \cdot \ve n|$ is called the \emph{spinning rate}, because it is the rate at which the particle spins around its symmetry axis.



\end{document}
